Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2022-12-22 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{methORF}). Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{methORF}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{methRNA}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible). The following equation describes the probability of gene gain:

\begin{equation}
P_\textit{gene-gain} = P_\textit{RNA-gain} \times P_\textit{ORF-stay} + P_\textit{ORF-gain} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
\label{genegaineq}
\end{equation}

Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}). Specifically:

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/geneGainLossComplete.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}. \autoref{genegaineq} describes total gene gain probability. Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$. Using the conditional probability of gene gain does not change these results.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$. To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms of gene gain and gene loss: $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. For example, the ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event. We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF. We found that for all ORF lengths, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}). This was also the case for a proto-gene containing 50 codons in its ORF and having different amounts of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).

We performed the same analysis on just the ORFs (and not the whole genes). That is, we calculated the ratio of logarithms of ORF gain and ORF loss. We found that for ORFs longer than 143 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.

@@

\begin{equation}
P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}
\label{rnafirsteq1}
\end{equation}

\begin{figure}[!t]
\centering
%\begin{subfigure}{0.48\textwidth}
%\centering
%\caption{One-step}
\includegraphics[scale=0.5]{Figures/whichfirstComplete.pdf}
%\end{subfigure}\hfill%
%\begin{subfigure}{0.48\textwidth}
%\centering
%\caption{Two-step}
%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
%\end{subfigure}
\caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first). \textbf{(A)} The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence, with a GC-content of 0.42 (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}). A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory. Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}). \textbf{(B)} Minimum ORF length (codons) that makes RNA-first trajectory more feasible (vertical axis) for different GC-content (horizontal axis). For a GC-content of 0.3 and 0.35, proto-genes of every analysed ORF length ($\geq33$ codons) preferentially emerge RNA-first.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}
We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first). In other words, the ORF-first hypothesis is more feasible. Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible. With the specific parameter values we used, it turns out that proto-genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). This minimum ORF length above which the RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content. For example, this number is 48 codons for a GC-content of 0.45 (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).

A more stringent definition of ORF-first hypothesis would describe a two-step process. In the first step, an ORF emerges in an untranscribed region of DNA but transcription does not emerge. In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}). Likewise, RNA-first hypothesis can be defined by a two-step probability. In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).

\begin{align}
P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
\end{align}

With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}\ \%$). However, we still find that proto-genes with shorter ORFs ($<$ 34 codons) have a slightly higher chance of emerging ORF first.

Overall, long ORFs are more likely to be lost before transcription emerges than short ORFs, and hence they preferentially emerge in existing transcripts. Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.

@@

\begin{figure}[!b]
\centering
\hrule
\vspace{1ex}
\includegraphics[scale=0.5]{Figures/pLossComplete.pdf}
\caption{ORF loss is more probable than transcription loss for proto-genes containing more than 33 codons. The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}. The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
\label{lossprob}
\end{figure}

@@

@@

\subsubsection{Definition of the probabilities}

Probability that a DNA region is transcribed by chance ($P_\textit{RNA}$), is the probability of finding a poly-A signal multiplied by the sum of probabilities of finding a TATA-box, and an Inr element.

\begin{equation}
P_\textit{RNA} = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr})
\label{eqrnaprob}
\end{equation}

Gain of transcription can happen via two mechanisms. In the first mechanism a promoter (TATA-box or Inr) already exists, and mutations cause a poly-A signal to emerge ($P_\textit{polyA-gain}$) but do not destroy the promoter ($P_\textit{TATA-stay}$,  $P_\textit{Inr-stay}$). In the second mechanism, a poly-A signal exists and the promoter emerges due to mutations ($P_\textit{TATA-gain}$,  $P_\textit{Inr-gain}$), while the poly-A signal remains undisrupted ($P_\textit{polyA-stay}$). Thus the total probability of transcription gain ($P_\textit{RNA-gain}$) would be the sum of the probabilities of the two mechanisms.

\begin{equation}
P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
\label{eqrnagain}
\end{equation}

@@

\begin{equation}
P_\textit{RNA-stay} = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
\label{eqrnastay}
\end{equation}

\begin{equation}
P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
\label{eqorfprob}
\end{equation}

\begin{align}
P_\textit{ORF-gain} & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
\label{eqorfgain}
\end{align}

\begin{equation}
P_\textit{ORF-loss} = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
\label{eqorfloss}
\end{equation}

\begin{equation}
P_\textit{ORF-stay} = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
\label{eqorfstay}
\end{equation}

>>>>

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}). Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible). We note that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present. This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence. This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}). Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed. Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively. We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:

\begin{equation}
P_\textit{gene-gain}^* = P_\textit{RNA-gain \vline\ ORF} \times P_\textit{ORF-stay} + P_\textit{ORF-gain \vline\ RNA} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
\label{genegaineq}
\end{equation}

\autoref{genegaineq} describes the total probability of gene gain that includes the probability that the gene was not previously present. To more precisely address the question of whether a gene that was absent in an ancestor but emerges in a descendant, we calculated the corresponding conditional probability of gene gain:

\begin{equation}
P_\textit{gene-gain} = P_\textit{gene-gain}^*/(1-P_\textit{ORF} - P_\textit{RNA})
\label{genegaineq2}
\end{equation}

We note that the difference between the total probability ($P_\textit{gene-gain}^*$) and the conditional probability ($P_\textit{gene-gain}$) of gene gain, is very small.

Next, we calculated the probability of gene loss, given the gene (transcription and ORF) is already present($P_\textit{gene-loss}$). It is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}). Specifically:

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/geneGainLossComplete2_new.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$. To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms (log-log ratio) of gene gain and gene loss for proto-genes with an ORF ranging from 30 -- 300 codons in length.

$$ \text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$$

For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event. We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF. We found that for ORF longer than 30 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}). To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons). We found that the effect of GC-content on the log-log ratio depends on the length of the ORF. Specifically, the log-log ratio steadily increases with GC-content ranging from 0.3 to 0.6, when the proto-gene contains an ORF with 30 codons. When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases. For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range). Overall, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.

We performed an analogous analysis on just the ORFs (and not the whole genes). That is, we calculated the ratio of logarithms of ORF gain and ORF loss. We found that for ORFs longer than 142 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.

@@

\begin{equation}
P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
\label{rnafirsteq1}
\end{equation}

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first. The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}). A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}. In both panels, the horizontal axis denotes the number of codons in the ORF. \textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and GC-content values at which ORF-first trajectory is more likely than RNA-first trajectory.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}

We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first). In other words, the ORF-first hypothesis is more feasible. Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible. We found that proto-genes with an ORF containing 60 codons or more, preferentially emerge RNA-first, whereas proto-genes with shorter ORFs emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). More specifically, the likelihood of the ORF-first trajectory relative to that of the RNA-first trajectory  decreases with increasing ORF length u a minimum value, and then increases. To understand if GC-content determines the evolutionary trajectory of \textit{de novo} emergence, we calculated the probabilities of the ORF-first and the RNA-first trajectory for different values of GC-content. We found that proto-genes with short ORF can emerge ORF-first at high GC-content. For example, the ORF-first trajectory is more likely at a GC-content of 0.6, for ORFs with 30 -- 36 codons (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).

A more stringent definition of the ORF-first hypothesis would describe a two-step process. In the first step, an ORF emerges in an untranscribed region of DNA but transcription does not emerge. In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}). Likewise, a RNA-first hypothesis can be defined by a two-step probability. In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).

\begin{align}
P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
\end{align}

With this stringent definition, proto-genes with every analysed ORF length (30 -- 300) preferentially emerge via the RNA-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}C}). Furthermore, the two-step RNA-first trajectory becomes increasingly more probable than the two-step ORF-first trajectory, with increasing ORF length.

Overall, long ORFs are likely to be lost before transcription emerges, and hence they preferentially emerge in existing transcripts. Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.

@@

\begin{figure}[!b]
\centering
\hrule
\vspace{1ex}
\includegraphics[scale=0.5]{Figures/pLossComplete2_new.pdf}
\caption{ORF loss in proto-genes is more probable than transcription loss. The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}. The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
\label{lossprob}
\end{figure}

@@

\begin{equation}
P_\textit{ORF}(k) = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
\label{eqorfprob}
\end{equation}

\begin{align}
P_\textit{ORF-gain}(k) & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
\label{eqorfgain}
\end{align}

\begin{equation}
P_\textit{ORF-loss}(k) = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
\label{eqorfloss}
\end{equation}

\begin{equation}
P_\textit{ORF-stay}(k) = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-gain})^{k-2}
\label{eqorfstay}
\end{equation}

@@

\subsubsection{Probability of transcription}

For a a DNA region of a length, to be transcribed it should have a promoter the beginning of the DNA region, and a poly-A signal at the end of it. The transcribed DNA region should also not have any poly-A signal sequences within it except the one at the end of it. A transcribed DNA of length $l$ nucleotides, contains $l-5$ 6-mers. Each of these 6-mers can be a poly-A signal with a probability, $P_\textit{polyA}$. Probability that none of these 6-mers is a poly-A signal is $(1-P_\textit{polyA})^{l-5}$. Since we focus on protein coding genes in this work, we describe the length of transcript in terms of the length of protein it can encode. Assuming that untranslated regions have no effect on protein synthesis, the length of a transcript in number of nucleotides should be at least three times the length of the encoded protein in number of amino acids. Thus the probability that a DNA region produces a transcript that can harbor an ORF with $k$ codons ($P_\textit{RNA}$) is:

\begin{equation}
P_\textit{RNA}(k) = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr}) (1-P_\textit{polyA})^{3k-5}
\label{eqrnaprob}
\end{equation}

\subsubsection{Probability of gaining transcription}

As we mentioned in the previous section, three requirements need to be met to produce a transcript of any given length. First, a promoter needs to be present. Second, a poly-A signal needs to be present at the end of the DNA region to be transcribed, and third, no poly-A signal should exist within this DNA region. Transcription can be gained if one of two of these requirements are already met and the third required feature emerges due to mutations. At the same time mutations do not destroy the existing features. It is possible that two or all the required features are missing and they emerge due to mutations. However such an event is highly improbable. Thus transcription gain can occur via three mechanisms.

\vspace{\baselineskip}
In the first mechanism:
\begin{itemize}
\item A promoter (TATA-box or Inr) exists at the beginning of a DNA region and remains intact ($P_\textit{TATA-stay}$, $P_\textit{Inr-stay}$)
\item No poly-A signal sequences are present within the DNA region and none emerges ($P_\textit{nopolyA-stay} = 1 - P_\textit{polyA} - P_\textit{polyA-gain}$)
\item A poly-A signal emerges at the end of the DNA region ($P_\textit{polyA-gain}$)
\end{itemize}

\vspace{\baselineskip}
In the second mechanism:
\begin{itemize}
\item A poly-A signal is present at the end of the DNA region remains intact 
\item No poly-A signal sequences are present within the DNA region and none emerges
\item A promoter emerges at the beginning of the DNA region ($P_\textit{TATA-gain}$, $P_\textit{Inr-gain}$)
\end{itemize}

\vspace{\baselineskip}
Finally, in the third mechanism:
\begin{itemize}
\item A promoter exists at the beginning of a DNA region and remains intact
\item A poly-A site is present at the end of the DNA region remains intact
\item One poly-A site that was present in the DNA region, is lost. ($P_\textit{polyA-loss}$)
\item At every other site in the DNA region, no poly-A signal exists and none emerges
\end{itemize}

The total gain probability ($P_\textit{RNA-gain}$) of a transcript that can harbor an ORF with $k$ codons  would be the sum of the probabilities of the three gain mechanisms:

\begin{align}
P_\textit{RNA-gain}(k)  = & \ (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber\\[1ex]
%
& + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber \\[1ex]
& + (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
\label{eqrnagain}
\end{align}

@@

\begin{equation}
P_\textit{RNA-stay}(k) = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
\label{eqrnastay}
\end{equation}

\subsection{Interdependence of transcription and ORF probabilities}

\label{rnaorfindependence}

All the four poly-A signal variants contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon. Since stop codons cannot exist inside an ORF, presence of an ORF reduces the possible number of poly-A signal sites in a DNA region. Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame. Thus, out of $3k-5$ positions for a 6-mer (length of a poly-A signal) in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}). Furthermore a poly-A signal cannot overlap with the start codon, thus reducing two more positions from all sites where a poly-A signal can exist (one position is already counted -- if a poly-A signal overlaps with the first codon in the second frame, then the second codon becomes a stop codon). Overall, the possible number poly-A sites on an ORF with $k$ codons is $2k-5$. Thus for a gene to emerge via gain of transcription, the DNA should not contain poly-A sites at these $2k-5$ positions. As a consequence, the probability of transcription gain for a specific transcript length, is higher when the DNA region contains an ORF than when it does not (\autoref{eqrnagain}). We have ignored the probability of poly-A signal overlapping with a stop codon because it makes a very small difference in the probability of RNA gain. This is so because at least one stop codon (\texttt{TAA}) allows a poly-A signal to overlap with it in three frames.

Conversely to the effect of an existing ORF on transcription gain, stop codons are less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region. Specifically, the probability that a 6-mer contains a stop codon in the third position is same as the probability of finding a stop codon ($P_\textit{stop}$). If poly-A signals are excluded from these 6-mers, then the probability of finding a stop codon is $P_\textit{stop} - P_\textit{poly-A}$. Conversely, lack of poly-A signals will increase the probability of amino acid coding codons and thereby their probability of mutating into a stop codon. Furthermore, the likelihood that an ORF is undisrupted my premature nonsense mutations, is lower when the DNA is already transcribed. Specifically:

$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} + P_\textit{poly-A})}{1- P_\textit{stop}}$$

Overall the evolutionary dynamics of transcription and ORF gain, are not independent of each other. That is, presence of one feature makes the gain of the other more likely. The dependence becomes more prominent with increasing number of codons in the ORF of the proto-gene.
